Parallel Adaptive Importance Sampling

### 2016-04-12 PAIS.tex
Updated problem B2

%Ideally the dimension of particles in a filter will not affect the convergence of the method. Unfortunately, the approximation of continuous density functions is computationally infeasible in high dimensions and the required ensemble size scales exponentially with the dimension of the problem\cite{silverman1986density,snyder2008obstacles}.

%\begin{figure}[htb]
%\begin{center}
%\includegraphics[width=\textwidth]{figures/ET_dim_scaling}
%\caption{Demonstration of the ETMC algorithm rapidly failing as the dimension increases. The relative error shown is that between the sampled mean and the true mean.}
%\label{fig:ET_dim_scaling}
%\end{center}
%\end{figure}

%Figure~\ref{fig:ET_dim_scaling} shows how quickly the ETMC algorithm loses its ability to accurately preserve the mean of the posterior distribution as the dimension increases. Clearly for a simple Gaussian problem, we can not resample in much more than 10 dimensions and hope to recover a representative sample, without hugely increasing the number of particles in the ensemble. As the cost of the ETMC is $\mathcal{O}(n^2)$, this does limit the dimension for which this approach can be used.

@@

Important sampling can be a very efficient method for sampling from a
probability distribution. A proposal density is chosen, from which we
can draw samples. Each sample is assigned a weight given by the
ratio of the target density and the proposal density at that
point. They are efficient when the proposal density is concentrated in
similar areas to the target density, and incredibly inefficient when
this is not the case. The aim of the PAIS is to use the states of a
set of parallel MCMC chains to construct a proposal distribution which
will be as close as possible to the target density. Given enough
processors, the states of all of these chains at one point
in time may be reasonably representative sample of the target density.

@@

%%% remove this paragraph?
Another related statistic, is the variance of the sample estimate of the first moment $\hat{\mu}$. Assuming that the algorithm is converging to the invariant distribution, convergence will be fastest when the variance of the estimate is minimised. This fact follows from the convergence rate of Monte Carlo algorithms, $\sigma/\sqrt{n}$. The proposal distribution which minimises the variance, $\sigma^2$, of $\hat{\mu}$, is the most desirable. This variance statistic often converges faster than the moment itself, and does not require any knowledge of the true value of the moment.

@@

\subsubsection{Numerical Implementation}\label{sec:Implementation P2}

The numerical implementation for the following simulations follow the same setup as described in Section~\ref{sec:Implementation P1}.

@@

\begin{table}[!htb]
    \begin{minipage}{.5\linewidth}
      \centering
        \begin{tabular}{|l|r|r|}
	\hline
	Algorithm							& $\delta^*_{\text{acc}}$	& $\delta^*_{\text{eff}}$ \\ \hline
	RWMH								& 1.3e-1					& - \\
	PAIS-RW								& -						& 5.4e-3\\
	\hline
	\end{tabular}
    \end{minipage}%
    \begin{minipage}{.5\linewidth}
      \centering
        \begin{tabular}{|l|r|r|r|}
	\hline
	Algorithm							& $\delta^*_{\text{acc}}$	& $\delta^*_{\text{eff}}$	& $\delta^*_{\text{L2}}$ \\ \hline
	RWMH								& 2.6e-2					& - 						& 9.1e-1\\
	PAIS-RW								& -						& 1.1e-3 					& 8.1e-3\\
	MALA								& 2.6e-2					& - 						& 1.3e-1\\
	PAIS-MALA							& -						& 3.7e-3 					& 8.1e-3\\
	\hline
	\end{tabular}
    \end{minipage}
	\vspace{1mm}
	\caption{Optimal values of $\delta$ for $B_1$ (left) and $B_2$ (right).}
	\label{table:BM_opt_delta}
\end{table}

Problem $B_2$ is much harder than $B_1$; transitions between the modes
are extremely unlikely for the standard pCNL algorithm. This means that we need to consider the
convergence on two levels; we should consider the algorithm's ability
to find all the modes, and to sample them thoroughly and in the correct proportions.

@@

From Table~\ref{table:BM_opt_delta} (right) we see that there is a large difference between the optimal value of $\delta$ for each regime in both RWMH and MALA, meaning that both will result in inefficient sampling. The PAIS-RW algorithm manages to sample the local detail and the large scale behaviour with similar values of $\delta^*$: a clear advantage to using this algorithm for this problem.

@@

\begin{figure}[htpb]
\begin{center}
\includegraphics[width=\textwidth]{"figures/BM1_L2"}
\caption{Error analysis for the PAIS-pCNL and pCNL algorithms for
  problem $B_1$. The solid blue line, and dashed red line compare the algorithms with fixed optimal scaling parameters, and the blue crossed and magenta crossed lines compare the adaptive algorithms. The setup is as described in Section~\ref{sec:Implementation P2} (2,3).}
\label{fig:BM1_L2}
\end{center}
\end{figure}

@@

For $B_2$ the algorithms are run with the global optimal value of $\delta^*$, and with the local optimal value of $\delta^*$. Figure~\ref{fig:BM2_L2} shows that the algorithms using the globally optimal $\delta^*$ convergence diagnostics converge slowly towards the true posterior after a long burn-in period, whereas the algorithms using the local optimal $\delta^*$ converge quickly, but get stuck in one mode meaning that the convergence rate flattens out.

\begin{figure}[h]
\begin{center}
\includegraphics[width=\textwidth]{"figures/BM2_L2"}
\caption{pCNL and PAIS-pCNL convergence statistics using locally and
  globally optimal $\delta$, for problem $B_2$. The setup is as described in Section~\ref{sec:Implementation P2} (2).}
\label{fig:BM2_L2}
\end{center}
\end{figure}

The adaptive algorithm as stated in Section~\ref{sec:adapt} is one way of combining the two regimes. Another method uses a small number of `scout' chains with a large $\delta$ to continually search out new modes. Other methods of mode searches are described in \cite{lan2013wormhole}, and the regeneration method is applicable\cite{nummelin1984general}.

Figure~\ref{fig:BM2_AL2} shows the success of the (A)PAIS-pCNL algorithms in converging to the posterior, compared to the (A)pCNL algorithms. We can see that using the pCNL algorithm for this problem would be infeasible.

\begin{figure}
\begin{center}
\includegraphics[width=\textwidth]{"figures/BM2_AL2"}
\caption{Convergence graphs for problem $B_2$, the (A)pCNL and (A)PAIS-pCNL algorithms have been run with optimal $\delta^*$ for the L2 error and the adaptive algorithm described in Section~\ref{sec:adapt}. The setup is as described in Section~\ref{sec:Implementation P2} (3).}
\label{fig:BM2_AL2}
\end{center}
\end{figure}

@@

\begin{figure}
\begin{center}
\includegraphics[width=\textwidth]{"figures/calc_saving"}
\caption{Illustration of calculating the number of PAIS-pCNL iterations required to reach a tolerance of $10^{-2}$ as a percentage of pCNL iterations. The relative L2 error graphs are from the problem in Section~\ref{sec:problem 1}.}
\label{fig:calc_saving}
\end{center}
\end{figure}

The graphs in the previous section clearly show that the PAIS-pCNL algorithm converges faster than the pCNL algorithm when both are parallelised with the same number of threads. We can calculate the number of iterations required to achieve a particular tolerance level in our solution for each algorithm and compare these to calculate a percentage saving. In Figure~\ref{fig:calc_saving} we demonstrate our calculation of the savings. The constants $c_1$ and $c_2$ are found by regressing through the data with a fixed exponent of $-1/2$ excluding the initial data points where the graph has not finished burning in.

A summary of the percentage of iterations required using the PAIS algorithm compared with the respective Metropolis-Hastings algorithms is given in Table~\ref{table:calc_savings}. The blank entries correspond to occasions when the MH algorithms haven't converged to the posterior distribution.

\begin{table}[!h]
\centering
\begin{tabular}{|l|l|r|r|}
\hline
		& & Low KL div & High KL div \\ \hline
	\multirow{4}{*}{Gaussian}	&RWMH & 14\% & 14\% \\
	&pCN & 33\% & - \\
	&MALA & 41\% & 42\% \\
	&pCNL & 62\% & 66\% \\ \hline
	\multirow{4}{*}{Bimodal}	&RWMH & 40\% & 44\% \\
	&pCN & 32\% & - \\
	&MALA & 36\% & 40\% \\
	&pCNL & 56\% & - \\ \hline
\end{tabular}
\caption{Iterations for the PAIS algorithms required to achieve a desired tolerance as a percentage of the number of iterations required by the respective Metropolis algorithms.}
\label{table:calc_savings}
\end{table}

>>>>

@@

Importance sampling can be a very efficient method for sampling from a
probability distribution. A proposal density is chosen, from which we
can draw samples. Each sample is assigned a weight given by the
ratio of the target density and the proposal density at that
point. They are efficient when the proposal density is concentrated in
similar areas to the target density, and incredibly inefficient when
this is not the case. The aim of the PAIS is to use the states of a
set of parallel MCMC chains to construct a proposal distribution which
will be as close as possible to the target density. Given enough
processors, the states of all of these chains at one point
in time may be reasonably representative sample of the target density.

@@

@@

@@

\begin{table}[!htb]
    \begin{minipage}{.5\linewidth}
      \centering
        \begin{tabular}{|l|r|r|}
	\hline
	Algorithm							& $\delta^*_{\text{acc}}$	& $\delta^*_{\text{eff}}$ \\ \hline
	RWMH								& 1.3e-1					& - \\
	PAIS-RW								& -						& 5.4e-3\\
	\hline
	\end{tabular}
    \end{minipage}%
    \begin{minipage}{.5\linewidth}
      \centering
        \begin{tabular}{|l|r|r|r|}
	\hline
	Algorithm							& $\delta^*_{\text{acc}}$	& $\delta^*_{\text{eff}}$	& $\delta^*_{\text{L2}}$ \\ \hline
	RWMH								& 2.6e-2					& - 						& 9.1e-1\\
	PAIS-RW								& -						& 1.3e-3 					& 8.1e-3\\
	\hline
	\end{tabular}
    \end{minipage}
	\vspace{1mm}
	\caption{Optimal values of $\delta$ for $B_1$ (left) and $B_2$ (right).}
	\label{table:BM_opt_delta}
\end{table}

It is relatively simple to find optimal scaling parameters for problem $B_1$. These values are given in Table~\ref{table:BM_opt_delta} (left). However problem $B_2$ is much harder; transitions between the modes are extremely unlikely for the standard RWMH algorithm. This means that we need to consider the
convergence on two levels; we should consider the algorithm's ability
to find all the modes, and also whether it can sample them in the correct proportions.

@@

From Table~\ref{table:BM_opt_delta} (right) we see that there is a large difference between the optimal value of $\delta$ for each regime in RWMH, and so will result in inefficient sampling. The PAIS-RW algorithm manages to sample the local detail and the large scale behaviour with similar values of $\delta^*$: a clear advantage to using this algorithm for this problem.

@@

\begin{figure}[htb]
\centering
\subfigure[Relative L2 error.]{\includegraphics[width=0.45\textwidth]{"figures/BM1_L2"}}
\subfigure[Absolute error in first moment.]{\includegraphics[width=0.45\textwidth]{"figures/BM1_moments"}}
\caption{Convergence of the PAIS-RW and RWMH algorithms for Problem $B_1$. Set up described in Section~\ref{sec:Implementation P1}.}
\label{fig:BM1_L2}
\end{figure}

@@

\begin{figure}[htb]
\centering
\subfigure[Relative L2 error using the optimal scaling parameters.]{\includegraphics[width=0.45\textwidth]{"figures/BM2_L2"}}
\subfigure[Comparing of the relative L2 error of the algorithms optimised by L2 minimisation with locally optimised PAIS-RW with a scout chain.]{\includegraphics[width=0.45\textwidth]{"figures/BM2_scout"}}
\caption{Convergence of the PAIS-RW and RWMH algorithms for Problem $B_2$. Set up described in Section~\ref{sec:Implementation P1}.}
\label{fig:BM2_L2}
\end{figure}

For $B_2$ the algorithms are run with the global optimal value of $\delta^*$, and with the local optimal value of $\delta^*$. Figure~\ref{fig:BM2_L2} (a) shows that the algorithms using the globally optimal $\delta^*$ convergence at the desired rate, whereas the algorithms optimised using the acceptance rate and effective sample size initially converge faster but at some point forget the location of one of the modes, causing the convergence to flatten out.

Using the PAIS algorithm we can minimise the impact of forgotten modes by constantly allowing the algorithm to search them out. Since we have parallel chains, we can run PAIS-RW with the majority of chains using the locally optimal value of $\delta$, and one or two chains with a larger scaling parameter. These chains with larger scaling parameters act both as `scouts' for new modes, and act to aid in the overdispersal of the proposal distribution. Figure~\ref{fig:BM2_L2} (b) shows the results of using 49 chains with the local optimal scaling parameter, and one chain with ten times the local optimal scaling parameter. We see that modes are not forgotten and the algorithm converges with the improvement we see from PAIS-RW in the other problems. Other methods of mode searches are described in \cite{lan2013wormhole}.

The adaptive algorithm can just as easily be applied to the PAIS algorithm with the `scout' chains described in the previous paragraph. Since we need two equally sized sub-ensembles, we will use two groups of 24 ensembles with the same proposal distributions, and each group will also have a `scout' chain with a scaling parameter ten times that of the rest of the group.

\begin{figure}[htb]
\begin{center}
\includegraphics[width=0.7\textwidth]{"figures/BM2_AL2"}
\caption{Convergence of the relative L2 error for problem $B_2$, comparing the globally optimised nonadaptive algorithms with the locally optimised adaptive algorithms. The setup is as described in Section~\ref{sec:Implementation P1} (3).}
\label{fig:BM2_AL2}
\end{center}
\end{figure}

Comparing the convergence of the adaptive algorithms against the nonadaptive algorithms in Figure~\ref{fig:BM2_AL2} shows that the algorithms behave as expected. The adaptive RWMH algorithm tuned using the acceptance rate converges at the same rate as the L2 optimised algorithm until the scaling parameter gets small where the algorithm forgets one of the modes. The adaptive PAIS-RW with scout chains tuned to the effective sample size converges at about the same rate as the locally optimised nonadaptive algorithm also with scouts.

@@

\begin{figure}
\begin{center}
\includegraphics[width=\textwidth]{"figures/calc_saving"}
\caption{Illustration of calculating the number of PAIS-pCNL iterations required to reach a tolerance of $10^{-2.5}$ as a percentage of pCNL iterations. The relative L2 error graphs are from the problem in Section~\ref{sec:problem 1}.}
\label{fig:calc_saving}
\end{center}
\end{figure}

The graphs in the previous section clearly show that the PAIS-RW algorithm converges faster than the RWMH algorithm when both are parallelised with the same ensemble size. We can calculate the number of iterations required to achieve a particular tolerance level in our solution for each algorithm and compare these to calculate a percentage saving. In Figure~\ref{fig:calc_saving} we demonstrate our calculation of the savings. The constants $c_1$ and $c_2$ are found by regressing through the data with a fixed exponent of $-1/2$ excluding the initial data points where the graph has not finished burning in.

A summary of the percentage of iterations required using the PAIS algorithm compared with the respective Metropolis-Hastings algorithms is given in Table~\ref{table:calc_savings}. The blank entries correspond to occasions when either the MH algorithm or PAIS algorithm hasn't converged to the posterior distribution.

\begin{table}[!h]
\centering
\begin{tabular}{|l|r|r|r|}
\hline
		& Gaussian & $B_1$ & $B_2$ \\ \hline
	RWMH & 10\% & 32\% & 12.6\% (scout) \\
	pCN & - & 32\% & - \\
	MALA & 42\% & 36\% & 40\% \\
	pCNL & 66\% & 56\% & - \\ \hline
\end{tabular}
\caption{Iterations for the PAIS algorithms required to achieve a desired tolerance as a percentage of the number of iterations required by the respective MH algorithms. The pCN and pCNL proposal distributions are taken from~\cite{cotter2013mcmc}.}
\label{table:calc_savings}
\end{table}
